---
title: "Reproducible Research: Peer Assessment 1"
author: "chhavi"
date: "Thursday, August 06, 2015"
output: html_document
---

Loading and preprocessing the data

```{r, echo=TRUE}

activity_raw <- read.csv("activity.csv", stringsAsFactors=FALSE)
```

Process/transform the data (if necessary) into a format suitable for analysis

```{r, echo=TRUE}
activity_raw$date <- as.POSIXct(activity_raw$date, format="%Y-%m-%d")


activity_raw <- data.frame(date=activity_raw$date, 
                           weekday=tolower(weekdays(activity_raw$date)), 
                           steps=activity_raw$steps, 
                           interval=activity_raw$interval)



activity_raw <- cbind(activity_raw, 
                      daytype=ifelse(activity_raw$weekday == "saturday" | 
                                       activity_raw$weekday == "sunday", "weekend", 
                                     "weekday"))

activity <- data.frame(date=activity_raw$date, 
                       weekday=activity_raw$weekday, 
                       daytype=activity_raw$daytype, 
                       interval=activity_raw$interval,
                       steps=activity_raw$steps)

```

We display the first few rows of the activity data frame:

```{r, echo=TRUE}
head(activity)
```

What is the mean total number of steps taken per day?
For this part of the assignment, you can ignore the missing values in the dataset.
Make a histogram of the total number of steps taken each day

```{r, echo=TRUE}
sum_data <- aggregate(activity$steps, by=list(activity$date), FUN=sum, na.rm=TRUE)
names(sum_data) <- c("date", "total")
```

We display the first few rows of the sum_data data frame:

```{r, echo=TRUE}
head(sum_data)
```

The histogram is given by the following lines of code:

```{r, echo=TRUE}
hist(sum_data$total, 
     breaks=seq(from=0, to=25000, by=2500),
     col="blue", 
     xlab="Total number of steps", 
     ylim=c(0, 20), 
     main="Histogram of the total number of steps taken each day\n(NA removed)")
```

Calculate and report the mean and median total number of steps taken per day....

below is the computation for   mean and median ................
```{r, echo=TRUE}
mean(sum_data$total)
median(sum_data$total)
```

These formulas gives a mean and median of 9354 and 10395 respectively.

......................................................................................................................

What is the average daily activity pattern?
Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r, echo=TRUE}
# Compute the means of steps accross all days for each interval
mean_data <- aggregate(activity$steps, 
                       by=list(activity$interval), 
                       FUN=mean, 
                       na.rm=TRUE)

# Rename the attributes
names(mean_data) <- c("interval", "mean")
```


first few rows of the mean_data data frame:

```{r, echo=TRUE}
head(mean_data)
```

The time serie plot is created by the following lines of code
```{r, echo=TRUE}
  plot(mean_data$interval, 
                            mean_data$mean, 
                            type="l", 
                            col="blue", 
                            lwd=2, 
                            xlab="Interval [minutes]", 
                            ylab="Average number of steps", 
                            main="Time-series of the average number of steps per intervals\n(NA removed)")

```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, echo=TRUE}
max_pos <- which(mean_data$mean == max(mean_data$mean))
max_interval <- mean_data[max_pos, 1]
```

.................................................................................................................

The 5-minute interval that contains the maximum of steps, on average across all days, is 835.

Inputing the missing values
Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NA’s)

```{r, echo=TRUE}
NA_count <- sum(is.na(activity$steps))
```


The number of NA’s is 2304.

.................................................................................................................


Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

```{r, echo=TRUE}
na_pos <- which(is.na(activity$steps))
mean_vec <- rep(mean(activity$steps, na.rm=TRUE), times=length(na_pos))
```

We use the strategy to remplace each NA value by the mean of the steps attribute.

Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r, echo=TRUE}
activity[na_pos, "steps"] <- mean_vec
```


We display the first few rows of the new activity data frame:
```{r, echo=TRUE}
head(activity)
```

.........................................................................................................


Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?



```{r, echo=TRUE}
sum_data <- aggregate(activity$steps, by=list(activity$date), FUN=sum)
names(sum_data) <- c("date", "total")
hist(sum_data$total, 
     breaks=seq(from=0, to=25000, by=2500),
     col="blue", 
     xlab="Total number of steps", 
     ylim=c(0, 30), 
     main="Histogram of the total number of steps taken each day\n(NA replaced by mean value)")
```


The mean and median are computed like

```{r, echo=TRUE}
mean(sum_data$total)
median(sum_data$total)
```


These formulas gives a mean and median of 10766 and 10766 respectively.

These values differ greatly from the estimates from the first part of the assignment. The impact of imputing the missing values is to have more data, hence to obtain a bigger mean and median value.


..................................................................................................................

Are there differences in activity patterns between weekdays and weekends?


For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

Create a new factor variable in the dataset with two levels - “weekdays” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r, echo=TRUE}
head(activity)
```

...............................................................................................................

Make a panel plot containing a time series plot (i.e. type = "l") of the 5- minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r, echo=TRUE}
library(lattice)

mean_data <- aggregate(activity$steps, 
                       by=list(activity$daytype, 
                               activity$weekday, activity$interval), mean)

names(mean_data) <- c("daytype", "weekday", "interval", "mean")

```

The time series plot take the following form:

```{r, echo=TRUE}
xyplot(mean ~ interval | daytype, mean_data, 
       type="l", 
       lwd=1, 
       xlab="Interval", 
       ylab="Number of steps", 
       layout=c(1,2))

```

....................................................................................................................
